---
title: "Class 11: Genome Informatics"
author: "Andrew Sue"
format: markdown_github
editor: visual
---

We are reading in the Mexican American Ancestry data for the SNPs associated with childhood asthma.

> Q1. What are those 4 candidate SNPs?

rs12936231, rs8067378, rs9303277, and rs7216389

> Q2. What three genes do these variants overlap or effect?

ZPBP2, IKZF3, GSDMB

> Q3. What is the location of rs8067378 and what are the different alleles for rs8067378?

Location: Chromosome 17:39895095

> Q4: Name at least 3 downstream genes for rs8067378?

Genes GSDMB, GSDMA and PSMD3.

```{r}
mxn_data <- read.csv("Asthma_genotypes.csv", row.names = 1)
mxn_data
```

> Q5. What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G\|G)?

14% G\|G phenotype.

```{r}
table(mxn_data$Genotype..forward.strand.) / nrow(mxn_data) * 100
```

> Q6. Back on the ENSEMBLE page, use the “search for a sample” field above to find the particular sample HG00109. This is a male from the GBR population group. What is the genotype for this sample?

The genotype of this sample is *G\|G*

We are reading in now the GBR dataset and finding the HG00109 sample.

```{r}
gbr <- read.csv("Asthma_genotypes_gbr.csv", row.names = 1)

table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100
```

To compare the gene expression of the entire population, we will take the population data.

```{r}
expr <- read.table("gene_expression.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

Lets make a summary plot (boxplot) of the data

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill = geno) + geom_boxplot(notch = TRUE)
```
